Typestate Inference in an Imperative Calculus

### 2013-01-06 formalism/paper.tex
removed unnecessary / unwritten proofs for remap, all typing proofs reviewed and fixed. principal typing proof still broken.

@@

\end{array}
\]
\caption{\label{fig:simreach} Definition of the simulation predecessor relation}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:unused %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{lem}
\label{lem:unused}
Given an arbitrary term $t$ such that 
$\typerule{\Gamma, x : T}{t}{T}{\Gamma', x : V}$ where $x$ is unused in
$t$, $T <: V$.
\end{lem}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:remapcv %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{itemize}
\item $U = \boolt$ or $\unitt$. By \remapunit or \remapbool, it follows
that $T = \remapapp{T}{U}{V}$. 
As $\boolt$ and $\unitt$ have no subtypes other than
themselves, it follows that $T' = T$. Trivially,
$\remapapp{T'}{U}{V} = \remapapp{T}{U}{V}$.

\item $U = \funt{U_i}{U_i'}{U'}$. By \remapfunc it follows that
$\remapapp{T'}{U}{V} = T' <: T = \remapapp{T}{U}{V}$.

\item $U = \ot{O_2}{S_2}$, with $T = \ot{O_1}{S_1}$ and $V = \ot{O_2}{S_4}$
and $\ot{O_1}{S_3} = \remapapp{T}{U}{V}$. Let $T' = \ot{O_3}{S_5}$,
and $\ot{O_3}{S_6} = \remapapp{T'}{U}{V}$. By definition,
$\overline{S_3} = \bigcup \{ \overline{S} \mid (\ot{O_1}{S_1}, \ot{O_2}{S_2}) \lesssim (\ot{O_1}{S}, \ot{O_2}{S_4}) \}$, and
$\overline{S_6} = \bigcup \{ \overline{S'} \mid (\ot{O_3}{S_5}, \ot{O_2}{S_2}) \lesssim (\ot{O_3}{S}, \ot{O_2}{S_4}) \}$.
In both cases, each $\overline{S}$ and $\overline{S'}$ extracted from
the path relation must be such that 
$(\ot{O_3}{S_5}, \ot{O_1}{S_1}) \lesssim (\ot{O_3}{S'}, \ot{O_1}{S})$.
Therefore, $\ot{O_3}{S_6} <: \ot{O_1}{S_3}$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:remap_in_contra %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{lem}
\label{lem:remap_in_contra}
Given types $T, U, U'$ and $V$ such that $U' <: U$ and that
both $\remapapp{T}{U'}{V}$ and $\remapapp{T}{U}{V}$ are defined,
$\remapapp{T}{U}{V} <: \remapapp{T}{U'}{V}$.
\end{lem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:remap_out_cv %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{lem}
\label{lem:remap_out_cv}
Given types $T, U, V$ and $V'$ such that $V' <: V$ and that
both $\remapapp{T}{U}{V}$ and $\remapapp{T}{U}{V'}$ are defined,
$\remapapp{T}{U}{V'} <: \remapapp{T}{U}{V}$.
\end{lem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:lesssim_same %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{proof}
TODO (coinductive?)
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:remapsame %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\item $T = \ot{O}{S_1}$. Let $\ot{O}{S_2} = \remapapp{T}{T}{T}$. By definition,
$\overline{S_2} = \bigcup \{ \overline{S} \mid (\ot{O}{S_1}, \ot{O_1}{S_1}) \lesssim (\ot{O_1}{S}, \ot{O_1}{S_1}) \}$. By \lemref{lem:lesssim_same}, it follows that
$\overline{S} = \overline{S_1}$ for all $\overline{S}$. Therefore,
$\overline{S_2} = \overline{S_1}$.
\end{itemize}
\end{proof}

\begin{lem}
\label{lem:remapeff}
Given types $T$, $U$ and $V$ such that $U <: V$.
It follows that $U <: \remapapp{U}{V}{V}$.
\end{lem}
\begin{proof}
By case analysis of the type $V$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:subst %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{lem}[Substitution]
\label{lem:subst}
If $\typerule{\Gamma}{t}{T}{\Gamma'}$, then
$\typerule
{\subst{x_i}{y_i}{\Gamma}}
{\subst{x_i}{y_i}{t}}
{T}
{\subst{x_i}{y_i}{\Gamma'}}$
where each $x_i$ and $y_i$ is distinct, and $\overline{x_i} \cap \overline{y_i} = \emptyset$.
\end{lem}
\begin{proof}
by induction on the structure of $t$.

If $t$ is some value $v$, then $\Gamma = \Gamma'$, therefore
$\subst{x_i}{y_i}{\Gamma} = \subst{x_i}{y_i}{\Gamma'}$
and substitution has no effect on the value, which has no specific requirements
of either $\Gamma$ or $\Gamma'$ other than that they are equal.
Therefore, $\typerule{\subst{x_i}{y_i}{\Gamma}}{\subst{x_i}{y_i}{v}}{T}{\subst{x_i}{y_i}{\Gamma'}}$.

If $t$ is a term, then it is of one of the following forms:

\begin{itemize}
\item $t = \lett{x}{t''}{t'}$. 
Substitution is defined on this term such that \\
$\subst{x_i}{y_i}{t} \equiv 
\lett{x}{( \subst{x_i}{y_i}{t''} )}{(\subst{x_j}{y_j}{t'})}$
, where $\overline{y_j} = \overline{y_i} - \{ x \}$.

By rule \tlet, it follows
that there exists $T', T'', \Gamma_1$ such that
$\typerule{\Gamma}{t''}{T''}{\Gamma_1}$ and that
$\typerule{\Gamma_1, x : T''}{t''}{T}{\Gamma', x : T'}$.
Additionally, $x \notin dom(\Gamma)$.

By induction, $t''$ can be substituted such that 
$\typerule
{\subst{x_i}{y_i}{\Gamma}}
{\subst{x_i}{y_i}{t''}}
{\subst{x_i}{y_i}{\Gamma_1}}
$.
Additionally, $t'$ can be substituted such that
$(\subst{x_j}{y_j}{\Gamma_1}) , x : T'' \triangleright \subst{x_j}{y_j}{t''} \triangleleft \{ (\overline{x_j / y_j} \} \Gamma') , x : T'$ as it is guaranteed that $x \notin \overline{y_j}$.
As $dom(\Gamma') = dom(\Gamma)$, and that $x \notin dom(\Gamma)$, it follows that $\subst{x_i}{y_i}{\Gamma_1} = (\subst{x_j}{y_j}{\Gamma_1})$ and
that $\overline{x_i / y_i} \} \Gamma' = \overline{x_j / y_j} \} \Gamma'$.

@@

\item $t = t_l ; t_r$.
Substitution is defined on this term such that
$\{ \overline{x_i / y_i} \} ( t_l ; t_r ) \equiv \subst{x_i}{y_i}{t_l} ; \subst{x_i}{y_i}{t_r}$.

By rule \tseq, there exists $\Gamma'', T'$ such that
$\typerule{\Gamma}{t_l}{T'}{\Gamma''}$ and
$\typerule{\Gamma''}{t_r}{T}{\Gamma'}$. By induction,
$\typerule{\subst{x_i}{y_i}{\Gamma}}{\subst{x_i}{y_i}{t_l}}{T'}{\subst{x_i}{y_i}{\Gamma''}}$ and
$\typerule{\subst{x_i}{y_i}{\Gamma''}}{\subst{x_i}{y_i}{t_r}}{T}{\subst{x_i}{y_i}{\Gamma'}}$.
Therefore, $t$ can by typed \tseq after the substitution.

\item $t = x_0 ( \overline{x_k} )$.
Substition is defined on this term such that
$\subst{x_i}{y_i}{x_0} ( \overline{x_k} ) \equiv y_0 ( \overline{y_k} )$
where for each $x \in \{ x_0, \overline{x_k} \}$,
$x = y_j \implies y = x_j$, otherwise $y$ is the original value.

For each substitution that occurs in the term, the variable must exist in
both the input and output contexts by rule \tfunc, and will be substituted
for the same name. Therefore, $t$ can be typed by \tfunc after the
substitution.

\item $t = x.m$.
Substition is defined on this term such that
$\subst{x_i}{y_i}{x.m} \equiv y.m$
where $y = x_j$ if $x = y_j$ for some $y_j \in \overline{y_i}$.
Otherwise, $y = x$.

Rule \tmethc requires that $x$ exist in both $\Gamma$ and $\Gamma'$ and
therefore any substitution of x in the term will be matched with the
same substitution in the contexts. Therefore, $t$ can be typed by
\tmethc after the substitution.

\item $t = \ift{t_c}{t_t}{t_f}$. Substitution is defined on this term such
that \\
$\subst{x_i}{y_i}{(\ift{t_c}{t_t}{t_f})} \equiv 
\ift{\substxy{t_c}}{\substxy{t_t}}{\substxy{t_f}}$.
By rule \tif there exists $\Gamma_1$, $\Gamma_2$, $\Gamma_3$, $T_t$, $T_f$
such that
$\typerule{\Gamma}{t_c}{\boolt}{\Gamma_1}$ and that
$\typerule{\Gamma_1}{t_t}{T_t}{\Gamma_2}$ and that
$\typerule{\Gamma_1}{t_f}{T_f}{\Gamma_3}$ where
$\Gamma' = \Gamma_2 \sqcup \Gamma_3$ and $T = T_t \sqcup T_f$. By induction
substitution can be performed on $t_c$, $t_t$ and $t_f$ such that
$\typerule{\substxy{\Gamma}}{\substxy{t_c}}{\boolt}{\substxy{\Gamma_1}}$ and that
$\typerule{\substxy{\Gamma_1}}{\substxy{t_t}}{T_t}{\substxy{\Gamma_2}}$ and
$\typerule{\substxy{\Gamma_1}}{\substxy{t_f}}{T_f}{\substxy{\Gamma_3}}$.
By the definition of substitution on contexts, 
$\substxy{\Gamma_2} \sqcup \substxy{\Gamma_3} = \substxy{\Gamma'}$. Therefore,
by rule \tif, 
$\typerule{\substxy{\Gamma}}{\substxy{\ift{t_c}{t_t}{t_f}}}{T}{\substxy{\Gamma'}}$.
\end{itemize}
\end{proof}

@@

\item $t = \funv{x_i}{T_i}{U_i}{t'}$. 
As $t$ is typeable, it must be the case that $t'$ is typeable also.
By induction, the term $t'$ has a principal typing
$(\Gamma, T, \Gamma')$. It follows that for each $i$, if $x_i \notin dom(\Gamma)$
then $x_i$ is not used in $t'$. Let $\overline{x_j} \subseteq \overline{x_i}$
be the set of such unused variables.
Let $\Gamma_2 = \Gamma, \overline{x_j : \top}$ and 
$\Gamma_2' = \Gamma', \overline{x_j : \top}$.
By \lemref{lem:weakening}, $(\Gamma_2, T, \Gamma_2')$ is a typing of $t'$.

If the effect $T_i \gg U_i$ is specified for parameter $x_i$ then it
must be the case that $T_i <: \Gamma_2(x_i)$ for each $i$ so that 
$t$ is typeable by rule \tfunc.
Let $\Gamma_3 = \overline{x_i : T_i'}$ such that:

\begin{itemize}
\item $T_i' = T_i$ if $T_i$ is specified
\item $T_i' = \Gamma_2(x_i)$ otherwise.
\end{itemize}

By repeated application of \lemref{lem:upgrade}, there exists a $\Gamma_3'$ and 
$T'$ such that
$(\Gamma_3,T',\Gamma_3')$ is a typing of $t'$, where $T' <: T$,
$dom(\Gamma_3') = dom(\Gamma_2')$ and 
$\forall x \in \Gamma_3' . \Gamma_3'(x) <: \Gamma_2'(x)$.

Let $U_i' = U_i$ if $U_i$ is specified or $U_i' = \Gamma_3'(x)$ otherwise.
It must be the case that $\Gamma_3'(x) <: U_i'$ for each $i$ so that
$t$ is typeable by rule \tfunc.
Let $\Gamma_4 = \overline{x_i : U_i}$.
By repeated application of rule \twiden, 
$(\Gamma_3,T',\Gamma_4)$ is a typing of $t'$.

>>>>

\newcommand{\pathrefl}{$\lesssim${\sc-refl}}
\newcommand{\pathadj}{$\lesssim${\sc-adj}}
\newcommand{\pathtrans}{$\lesssim${\sc-trans}}
\newcommand{\pathbranch}{$\lesssim${\sc-branch}}

@@

\end{array}
\]
\caption{\label{fig:simreach} Definition of the path relation}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:unused %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% last revised and checked: 2013/01/04

\begin{lem}
\label{lem:unused}
Given an arbitrary term $t$ such that 
$\typerule{\Gamma, x : T}{t}{U}{\Gamma', x : V}$ where $x$ is unused in
$t$, $T <: V$.
\end{lem}

\begin{proof}
Straightforward observation from the typing rules.
\end{proof}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:remapcv %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% revised and checked 2013/01/04

@@

\begin{itemize}
\item $U = \boolt$ or $\unitt$. By \remapunit or \remapbool, it follows
that $T = \remapapp{T}{U}{V}$. 
As $\boolt$ and $\unitt$ have no subtypes other than
themselves, it follows that $T' = T$. By \subrefl,
$\remapapp{T'}{U}{V} <: \remapapp{T}{U}{V}$.

\item $U = \funt{U_i}{U_i'}{U'}$. By \remapfunc it follows that
$\remapapp{T'}{U}{V} = T'$ and $\remapapp{T}{U}{V} = T$, therefore
$\remapapp{T'}{U}{V} <: \remapapp{T}{U}{V}$.

\item $U = \ot{O_1}{S_1}$. By \remapobj it follows that 
$V = \ot{O_1}{S_2}$ and
$T = \ot{O_2}{S_3}$ with
$\remapapp{T}{U}{V} = \ot{O_2}{S_4}$. Similarly,
$T' = \ot{O_3}{S_5}$
with $\remapapp{T'}{U}{V} = \ot{O_3}{S_6}$.

By definition,
$\overline{S_4} = \bigcup \{ \overline{S} \mid (\ot{O_2}{S_3}, \ot{O_1}{S_1}) \lesssim (\ot{O_2}{S}, \ot{O_1}{S_2}) \}$, and
$\overline{S_6} = \bigcup \{ \overline{S'} \mid (\ot{O_3}{S_5}, \ot{O_1}{S_1}) \lesssim (\ot{O_3}{S'}, \ot{O_1}{S_2}) \}$.
In both cases, each $\overline{S}$ and $\overline{S'}$ extracted from
the path relation must be such that 
$(\ot{O_3}{S_5}, \ot{O_2}{S_3}) \lesssim (\ot{O_3}{S'}, \ot{O_2}{S})$
as $T' <: T$.
Therefore, $\ot{O_3}{S_6} <: \ot{O_2}{S_4}$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:lesssim_same %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% revised and checked: 2013/01/04

@@

\begin{proof}
by induction on the path relation. Assume 
$(\ot{O}{S}, \ot{O}{S}) \lesssim (\ot{O}{S'}, \ot{O}{S''})$. It follows that
this will have been derived by one of the following rules:

\item \pathrefl. Directly, $\overline{S} = \overline{S'} = \overline{S''}$.

\item \pathadj. It follows that there exists a method $m$ such that
$m : T \Rightarrow \overline{S'} \in \ot{O}{S}$. There can be no other
target state set for this method, therefore $\overline{S'} = \overline{S''}$.

\item \pathtrans. It follows that there exists state sets $\overline{S_1}$
and $\overline{S_2}$ such that
$(\ot{O}{S}, \ot{O}{S}) \lesssim (\ot{O}{S_1}, \ot{O}{S_2})$ and
$(\ot{O}{S_1}, \ot{O}{S_2}) \lesssim (\ot{O}{S'}, \ot{O}{S''})$. By induction,
it follows that $\overline{S_1} = \overline{S_2}$. Therefore, also by induction,
$\overline{S'} = \overline{S''}$.

\item \pathbranch. It follows that there exists state sets
$\overline{S_1}$, $\overline{S_2}$, $\overline{S_3}$, $\overline{S_4}$ such
that 
$(\ot{O}{S}, \ot{O}{S}) \lesssim (\ot{O}{S_1}, \ot{O}{S_2})$ and
$(\ot{O}{S}, \ot{O}{S}) \lesssim (\ot{O}{S_3}, \ot{O}{S_4})$ where
$\overline{S'} = \overline{S_1} \cup \overline{S_3}$ and
$\overline{S''} = \overline{S_2} \cup \overline{S_4}$.
By induction, $\overline{S_1} = \overline{S_2}$ and 
$\overline{S_3} = \overline{S_4}$. Therefore, $\overline{S'} = \overline{S''}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:remapsame %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% revised and checked: 2013/01/04

@@

\item $T = \ot{O}{S_1}$. Let $\ot{O}{S_2} = \remapapp{T}{T}{T}$. By definition,
$\overline{S_2} = \bigcup \{ \overline{S} \mid (\ot{O}{S_1}, \ot{O_1}{S_1}) \lesssim (\ot{O_1}{S}, \ot{O_1}{S_1}) \}$. By \lemref{lem:lesssim_same}, it follows that
$\overline{S} = \overline{S_1}$ for all $\overline{S}$. Therefore,
$\overline{S_2} = \overline{S_1}$, meaning
$T = \remapapp{T}{T}{T}$.
\end{itemize}
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:remapsame %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% revised and checked 2013/01/04

\begin{lem}
\label{lem:remapeff}
Let $U <: V$. It follows that $U <: \remapapp{U}{V}{V}$.
\end{lem}
\begin{proof}
By case analysis of the type $V$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:subst %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% revised and checked: 2013/01/05

\begin{lem}[Substitution]
\label{lem:subst}
If $\typerule{\Gamma}{t}{T}{\Gamma'}$, then
$\typerule
{\subst{x_i}{y_i}{\Gamma}}
{\subst{x_i}{y_i}{t}}
{T}
{\subst{x_i}{y_i}{\Gamma'}}$
where each $x_i$ and $y_i$ is distinct, and $\overline{x_i} \cap \overline{y_i} = \emptyset$.
\end{lem}
\begin{proof}
by induction on the typing derivation of $t$. $\typerule{\Gamma}{t}{T}{\Gamma'}$
must have been derived by one of the following rules:

\item \tunit, \ttrue, \tfalse, \tobj or \tfundef. For each of these rules,
$\Gamma = \Gamma'$. Therefore, $\subst{x_i}{y_i}{\Gamma} = \subst{x_i}{y_i}{\Gamma'}$.
and substitution has no effect on the value. The rule used for derivation
has no specific requirements of either $\Gamma$ or $\Gamma'$ other than that 
they are equal.
Therefore, $\typerule{\subst{x_i}{y_i}{\Gamma}}{\subst{x_i}{y_i}{v}}{T}{\subst{x_i}{y_i}{\Gamma'}}$.

\item \tlet. It follows that $t = \lett{x}{t_v}{t_b}$. 
Substitution is defined on this term such that \\
$\subst{x_i}{y_i}{t} \equiv 
\lett{x}{( \subst{x_i}{y_i}{t_v} )}{(\subst{x_j}{y_j}{t_b})}$
, where $\overline{y_j} = \overline{y_i} - \{ x \}$.

By rule \tlet, it follows
that there exists $T', T'', \Gamma_1$ such that
$\typerule{\Gamma}{t_v}{T''}{\Gamma_1}$ and that
$\typerule{\Gamma_1, x : T''}{t_b}{T}{\Gamma', x : T'}$.
Additionally, $x \notin dom(\Gamma)$.

By induction, $t_v$ can be substituted such that 
$\typerule
{\subst{x_i}{y_i}{\Gamma}}
{\subst{x_i}{y_i}{t_v}}
{\subst{x_i}{y_i}{\Gamma_1}}
$.
Additionally, $t_b$ can be substituted such that
$(\subst{x_j}{y_j}{\Gamma_1}) , x : T'' \triangleright \subst{x_j}{y_j}{t_b} \triangleleft \{ (\overline{x_j / y_j} \} \Gamma') , x : T'$ as it is guaranteed that $x \notin \overline{y_j}$.
As $dom(\Gamma') = dom(\Gamma)$, and that $x \notin dom(\Gamma)$, it follows that $\subst{x_i}{y_i}{\Gamma_1} = (\subst{x_j}{y_j}{\Gamma_1})$ and
that $\overline{x_i / y_i} \} \Gamma' = \overline{x_j / y_j} \} \Gamma'$.

@@

\item \tseq. It follows that $t = t_l ; t_r$,
and there exists $\Gamma'', T'$ such that
$\typerule{\Gamma}{t_l}{T'}{\Gamma''}$ and
$\typerule{\Gamma''}{t_r}{T}{\Gamma'}$.
Substitution is defined on this term such that
$\subst{x_i}{y_i}{(t_l ; t_r)} \equiv \subst{x_i}{y_i}{t_l} ; \subst{x_i}{y_i}{t_r}$.

By induction, $t_l$ and $t_r$ can be typed such that
$\typerule{\subst{x_i}{y_i}{\Gamma}}{\subst{x_i}{y_i}{t_l}}{T'}{\subst{x_i}{y_i}{\Gamma''}}$ and
$\typerule{\subst{x_i}{y_i}{\Gamma''}}{\subst{x_i}{y_i}{t_r}}{T}{\subst{x_i}{y_i}{\Gamma'}}$.
Therefore, $t$ can by typed using \tseq such that

$\typerule{\subst{x_i}{y_i}{\Gamma}}{\subst{x_i}{y_i}{t}}{T}{\subst{x_i}{y_i}{\Gamma'}}$.

\item \tfunc. It follows that $t = z ( \overline{z_k} )$ and
that $\Gamma = \Gamma_1, \overline{z_k : T_k}$,
$\Gamma' = \Gamma_1, \overline{z_k : T_k'}$
where $\Gamma_1(z) = \funt{U_k}{V_k}{T}$ with for each $k$,
$T_k <: U_k$ and $T_k' = \remapapp{T_k}{U_k}{V_k}$. Each
variable in $\overline{z_k} \cup \{ z \}$ is distinct.

Substition is defined on this term such that
$\subst{x_i}{y_i}{z ( \overline{z_k} )} \equiv z' ( \overline{z_k'} )$
where $z' = z$ if $z \notin \overline{y_i}$ or $z = x_i$ for the $x_i$ such
that $z = y_i$, and similarly for all $z_i$.

The uniqueness of each substitution pair ensures
that no variable will be passed twice as part of the function call.
Therefore, the conditions of \tfunc are satisfied such that

$\typerule{\substxy{\Gamma}}{\substxy{t}}{T}{\substxy{\Gamma'}}$.

\item \tmethc. It follows that $t = x.m$ with
$\Gamma = \Gamma_1, x : \ot{O}{S}$ and
$\Gamma' = \Gamma_1, x : \ot{O}{S'}$ such
that $m : T \Rightarrow \overline{S'} \in \ot{O}{S}$.

Substition is defined on this term such that
$\subst{x_i}{y_i}{x.m} \equiv z.m$
where $z = x_j$ if $x = y_j$ for some $y_j \in \overline{y_i}$.
Otherwise, $z = x$.

The conditions of \tmethc are satisfied such that
$\typerule{\substxy{\Gamma}}{\substxy{t}}{T}{\substxy{\Gamma'}}$.

\item \tif. It follows that $t = \ift{t_c}{t_t}{t_f}$,
where there exists $\Gamma_1$, $\Gamma_2$, $\Gamma_3$, $T_t$, $T_f$
such that
$\typerule{\Gamma}{t_c}{\boolt}{\Gamma_1}$ and that
$\typerule{\Gamma_1}{t_t}{T_t}{\Gamma_2}$ and that
$\typerule{\Gamma_1}{t_f}{T_f}{\Gamma_3}$ where
$\Gamma' = \Gamma_2 \sqcup \Gamma_3$ and $T = T_t \sqcup T_f$.

Substitution is defined on this term such
that \\
$\subst{x_i}{y_i}{(\ift{t_c}{t_t}{t_f})} \equiv 
\ift{\substxy{t_c}}{\substxy{t_t}}{\substxy{t_f}}$.

By induction
substitution can be performed on $t_c$, $t_t$ and $t_f$ such that
$\typerule{\substxy{\Gamma}}{\substxy{t_c}}{\boolt}{\substxy{\Gamma_1}}$ and that
$\typerule{\substxy{\Gamma_1}}{\substxy{t_t}}{T_t}{\substxy{\Gamma_2}}$ and
$\typerule{\substxy{\Gamma_1}}{\substxy{t_f}}{T_f}{\substxy{\Gamma_3}}$.
By the definition of substitution on contexts, 
$\substxy{\Gamma_2} \sqcup \substxy{\Gamma_3} = \substxy{\Gamma'}$.

The conditions of the rule \tif are satisfied such that
$\typerule{\substxy{\Gamma}}{\substxy{\ift{t_c}{t_t}{t_f}}}{T}{\substxy{\Gamma'}}$.

\item \tdrop. It follows that there exists a $\Gamma''$ such that
$\typerule{\Gamma}{t}{T}{\Gamma', \Gamma''}$. By induction,
$\typerule{\substxy{\Gamma}}{\substxy{t}}{T}{\substxy{\Gamma', \Gamma''}}$.
By the definition of substitution on contexts,
$\substxy{\Gamma', \Gamma''} = \substxy{\Gamma'}, \substxy{\Gamma''}$.

Therefore by rule \tdrop,
$\typerule{\substxy{\Gamma}}{\substxy{t}}{T}{\substxy{\Gamma'}}$.

\item \tsub. It follows that
$\typerule{\Gamma}{t}{T'}{\Gamma'}$ for some $T' <: T$. By induction,
$\typerule{\substxy{\Gamma}}{\substxy{t}}{T'}{\substxy{\Gamma'}}$.
By rule \tsub,
$\typerule{\substxy{\Gamma}}{\substxy{t}}{T}{\substxy{\Gamma'}}$.

\item \twiden. It follows that $\Gamma' = \Gamma_1, x : U$
and there exists a $U' <: U$ such that
$\typerule{\Gamma}{t}{T'}{\Gamma_1, x : U'}$. 
By induction,
$\typerule{\substxy{\Gamma}}{\substxy{t}}{T}{\substxy{\Gamma_1, x : U'}}$.
By rule \twiden,
$\typerule{\substxy{\Gamma}}{\substxy{t}}{T}{\substxy{\Gamma_1, x : U}}$.

\item \tnarrow. It follows that $\Gamma = \Gamma_1, x : U'$
and there exists a $U$ such that $U' <: U$ and
$\typerule{\Gamma_1, x : U}{t}{T'}{\Gamma'}$.
By induction,
$\typerule{\substxy{\Gamma_1, x : U}}{\substxy{t}}{T}{\substxy{\Gamma'}}$.
By rule \tnarrow,
$\typerule{\substxy{\Gamma_1, x : U'}}{\substxy{t}}{T}{\substxy{\Gamma'}}$

@@

\item $t = \funv{x_i}{T_i}{U_i}{t'}$. 
As $t$ is typeable, it must be the case that $t'$ is typeable also. Trivially
$t'$ will contain no function calls as it is part of $t$.
By induction, the term $t'$ has a principal typing
$(\Gamma, T, \Gamma')$. In order for $t$ to be typeable, it must be the case
that $dom(\Gamma) \subseteq \overline{x_i}$ --- the function body cannot use
any variables which are not explicitly passed as a parameter.

For each $i$, if $x_i \notin dom(\Gamma)$
then $x_i$ is not used in $t'$. Let $\overline{x_j} \subseteq \overline{x_i}$
be the set of such unused variables.
Let $\Gamma_2 = \Gamma, \overline{x_j : \top}$ and 
$\Gamma_2' = \Gamma', \overline{x_j : \top}$.
By \lemref{lem:weakening}, $(\Gamma_2, T, \Gamma_2')$ is a typing of $t'$.

The effect $T_i \gg U_i$ for parameter $x_i$ must be such that 
$T_i <: \Gamma_2(x_i)$ and $\Gamma_2'(x_i) <: U_i$ in order for
$t$ to be typeable. 
Trivially, $(\emptyset, \funt{T_i}{U_i}{T}, \emptyset)$ is the principal
typing of $T$.

It is worth noting that the explicit effects for each parameter are not
actually necessary, and can be extracted directly from $\Gamma_2$
and $\Gamma_2'$. By choosing $T_i' = \Gamma_2(x_i)$ and $U_i' = \Gamma_2'(x_i)$,
$(\emptyset, \funt{T_i'}{U_i'}{T}, \emptyset)$ is a typing of the modified
function value $\lambda (\overline{x_i}) . t'$, which can safely omit the
effect types for each $x_i$.

%! extended proof of construction of principal types when some annotations
%! provided, and some are missing
%
% so that 
% $t$ is typeable by rule \tfunc.
% Let $\Gamma_3 = \overline{x_i : T_i}$.

% By repeated application of \lemref{lem:upgrade}, there exists a $\Gamma_3'$ and 
% $T'$ such that
% $(\Gamma_3,T',\Gamma_3')$ is a typing of $t'$, where $T' <: T$,
% $dom(\Gamma_3') = dom(\Gamma_2')$ and 
% $\forall x \in \Gamma_3' . \Gamma_3'(x) <: \Gamma_2'(x)$.
